Allowing Stable Groups of Prosumers to Enter the Market through Proper Aggregation

### 2014-09-05 Article.tex
update

% author names and affiliations
% use a multiple column layout for up to three different
% affiliations
\author{\IEEEauthorblockN{Nicolas Gensollen}
\IEEEauthorblockA{Institut Mines Telecom\\
Email: http://www.michaelshell.org/contact.html}
\and
\IEEEauthorblockN{Vincent Gauthier}
\IEEEauthorblockA{Twentieth Century Fox\\
Springfield, USA\\
Email: homer@thesimpsons.com}
\and
\IEEEauthorblockN{Michel Marot\\ and Monique Becker}
\IEEEauthorblockA{Starfleet Academy\\
San Francisco, California 96678-2391\\
Telephone: (800) 555--1212\\
Fax: (888) 555--1212}}

@@

\begin{abstract}
%\boldmath
There is a growing interest for coalitional concepts in the smart grid, spanning multiple objectives such as power losses reduction, benefits maximization or market stability... In this paper, we study coalitions of prosumers (agents that produce but also consume depending on time) which aim at selling energy to the grid. We reckon that the grid is able to specify two requirements (stability and minimum production) that enable a coalition to enter the market. Based on probability distributions of units available power, coalitions which are able to achieve low standard deviations are considered as stable coalitions. As it seems intuitively understandable that maintaining stable coalitions would be less expensive in terms of communication load than highly versatible ones, our objective consists in forming coalitions that satisfy grid requirements both in terms of stability and minimum production. Our definition of stability being implicitely linked to correlation between agents timeseries (see section \ref{sec:model}), we organize them over a correlation graph and use a variation of clique percolation to construct valid coalitions in reasonable time.
\end{abstract}
% IEEEtran.cls defaults to using nonbold math in the Abstract.
% This preserves the distinction between vectors and scalars. However,
% if the conference you are submitting to favors bold math in the abstract,
% then you can use LaTeX's standard command \boldmath at the very start
% of the abstract to achieve this. Many IEEE journals/conferences frown on
% math in the abstract anyway.

@@

In this paper, our focus is on how coalitions of prosumers, aiming at selling energy to the grid, should be formed in order to support it efficiently. Namely, improving statistically the production stability while decreasing the communication needs for staying in a stable state. The idea is that, by using prediction techniques, coalitions can propose contract values to the grid, enabling it to schedule production on its side accordingly. Actually, contract values are particularly relevant in power exchanges contexts where energy is traded based on predictions (day-ahead estimations for instance). It seems obvious that, in such systems, participants should try to minimize prediction errors in order to maintain a stable state of grid operation, and that some penalty rules should be applied to ensure that coalitions are willing to report contract values correctly. Furthermore, as the coalition productions are supposed to come from (uncertain) renewables, the notion of reliability comes into play with a visible tradeoff between contract values worth and reliability.

In this paper, our concern is on how prosumers should be aggregated in order to optimize this tradeoff, that is, how coalitions should be formed with the intention of announcing “high contract values with high reliability”. We thus aim at:

@@

On a narrower scale, the authors of \cite{Kota2011} study the formation of virtual power plants (VPP) composed of multiple self-interested DER. On the grid side, two requirements for the formation of virtual power plants are considered, namely the reliability of supply and the minimization of entities the grid has to deal with. From this, \cite{Kota2011} builds a pricing mechanism that encourages VPP to report true estimates of their aggregated production and penalizes prediction errors. A redistribution scheme of the VPP to the DER is also constructed such that the payoff allocation lies in the core of the game, meaning that no DER has an incentive to leave the coalition.

@@

The setting is thus similar with some fiancial studies on stock exchanges, where researchers tried to find relevant clusters of stocks based on their daily prices variations. The problem of clustering is usually approached by means of a similarity measure and completed by a clustering technique such as K-means or hierarchical clustering. Nevertheless, in \cite{Mantegna1999} Mantegna introduced another approach where stocks, represented by the timeseries of their daily log returns, are organized in a graph such that stocks (vertices) exhibiting similar price fluctuation patterns are close to each other. This closeness notion is formalized with a similarity measure based on Pearson correlation coefficient ($ d_{ij} = \dfrac{1}{2}\sqrt{2(1-\rho_{ij})} $ but sometimes also $ d_{ij} = 1 - \rho_{ij}^{2} $ ) that enables to weight the edges of the graph. Because there is a distance between every two nodes, \cite{Mantegna1999} obtains a complete graph where precious information is flooded. In order to discard unsignificant information, \cite{Mantegna1999} used a minimum spanning tree oriented technique and achieves a hierarchical clustering of the stocks. Since then, several other filtering techniques have been proposed, such as k-nearest graphs, or $ \epsilon $-graphs, that relax the tree assumption \cite{Onnela2004}.

@@

Presented in this way, the timeseries clustering task seems very close to graph community detection. Communities in networks are indeed often seen as groups of nodes exhibiting high internal densities of links as well as a low density across communities \cite{Newmanb}. Although several techniques exists (based on different graph properties : modularity \cite{Newmanb}, edge betweeness \cite{Girvan2002}, spectrum of Laplacian \cite{Newman}...), the clique percolation algorithm \cite{Lancichinetti} uses directly this observation and the idea behind it is actually quite simple. It starts indeed by searching for cliques (a complete subgraph) in the given graph as potential community cores. Because other nodes can naturally belong to a community with a less strong connectivity requirement, the algorithm uses a fitness function that quantifies the quality (according to some criterions) of a community structure inside the graph. Expansed cliques alternatively look in their respective neighborhood, select the node that yields the best increase in fitness, and finally incorporate it. The percolation algorithm stops when seeds cease to grow or when all nodes are affected to at least one seed. At this point, some coalitions may be partially overlapping, meaning that a node can belong to multiple communities. Detection of overlapping communities is actually a very active field of research, especially in social networks where persons often belongs to several communities (family, friends, work colleagues...).

@@

A major concern while designing our model was its ability to translate correctly realistic patterns (consumption and production) as well as realistic geographical correlations between patterns. Largely because of these reasons (but also because they are easily, and sometimes freely, available online) we chose to use meteorological traces as inputs. In this paper we used namely both french data from 2006 to 2012 sampled every 3 hours \cite{Infoclimat} (similar data for the United States covering year 2010 can also be found at \cite{NCDC}). The weather stations already offer a quite good sampling of a given territory, and, with the growth of personal weather stations constantly updating data bases, the mesh becomes finer and finer. As shown in the first part of figure 1, the first step consists in discretizing the studied zone around well chosen weather stations and gathering traces for these stations. In this paper, we consider that prosumers can only produce through wind turbines and photovoltaic panels (PV). On the other hand, we built their consumption patterns according to two major cycles :
\begin{itemize}
\item \textbf{Daily cycles} : Consumption is low during night, and higher during the day with two picks in the morning and evening. Some noise is added so that prosumers have similar but not identical cycles.
\item \textbf{Seasonal cycles} : Consumption is higher in the winter because of heating and low in the summer (we did not consider air conditioning in this paper). Temperature traces were the principal ingredient for modeling these cycles.
\end{itemize}

@@

At this point, modeling agents consists in fixing a few parameters (most of the time drawn from random distributions) such as the geographical position, the number of wind turbines, PV, appliances, the temperature of confort, and so on... The objective is that, depending on the weather of her zone, the DER and appliances she owns, and the way she decides to heat her home, a given prosumer is able to compute her production/consumption at any time. We thus used simple and known power curve models for linking weather variables to output power, such that a wind turbine (or a PV for instance) takes a wind speed (a solar irradiance) as input and gives an instantaneous output power \cite{Kota2011} \cite{windturbinemodel}. Any agent is thus able to compute at any time (within the traces bounds) how much she produces and consumes.

@@

The reliability stipulates that for any coalition S willing to join the market, the probability of S value (at any instant t) being below its contract value should at most be $ \phi $. That is, $ \forall S,\ \forall t,\ Pr[P_{S}(t) \leq P_{S}^{CRCT} ] \leq \phi $. Furthermore, for consistency, we restrict $ \phi $ to small values ( $ \phi << 0.5 $). The minimum contract value under $ \phi $ states that only coalitions with higher contract values ($ P_{S}^{CRCT} \geq P_{\phi}^{MIN} $) will be accepted.

At this point, if a coalition S whishes to join the market, it has to choose a contract value that fulfill the two conditions. For simplification, we consider that coalitions will always apply the same economically consistent strategy of announcing the highest possible contract value that obeys the reliability rule (a value we denote by $ P_{\phi}(S) $). If $ P_{\phi}(S) \geq P_{\phi}^{MIN} $, meaning that it also obeys the minimum contract value rule, then S annonces this value on the market : $ P_{S}^{CRCT} = P_{\phi}(S) $. Otherwise, coalition S is not able to enter.
Basically, a coalition S is valid if and only if :
\[ \left\{ \begin{array}{lll}
		\forall t,\ Pr[ P_{S}(t) \leq P_{\phi}(S)] \leq \phi\ \textit{{\scriptsize (reliability rule)}} \\
		and\ P_{\phi}(S) \geq P_{\phi}^{MIN}\ \textit{{\scriptsize (min value rule)}}

\end{array} \right. \]

We now choose a very simple utility function that derives directly from the above remarks. If a coalition cannot provide a valid contract value, it receives naturally a utility of zero. Furthermore, it seems obvious that the utility should increase with the contract value. The $ 1/|S| $ term indicates that we favorise small coalitions, mainly because they are easier to maintain in terms of communications.

\[ \mathcal{U}_{\phi,\ P_{\phi}^{MIN}}(S) = \mathbf{1}_{\textit{S\ valid}} \dfrac{P_{\phi}(S)}{|S|} \]

@@

\[ \left\{ \begin{array}{lll}
		\mu_{ij} = \mu_{i} + \mu_{j} \\
		\sigma_{ij} = \sqrt{\sigma_{i}^{2} + \sigma_{j}^{2} + \rho_{ij}\sigma_{i}\sigma_{j}}
\end{array} \right. \]

@@

\[ \dfrac{1}{2} \left[ 1+ erf \left( \dfrac{P_{\phi}(ij) - \mu_{ij}}{\sigma_{ij}\sqrt{2}} \right) \right] \leq \phi \]

The strategy of $ \{ij\} $ consists in maximizing its contract value as long as it respects this inequality, e.g to annonce $ P_{\phi}(ij) = \mu_{ij} - \sigma_{ij}\sqrt{2}erf^{-1}(1-2 \phi ) $. It thus appears (as it was intuitively understandable) that, for equivalent sizes, coalitions with low relative standard deviations ( $ \sigma_{ij} / \mu_{ij} $ ) are able to announce higher contract values.

@@

The model we used in order to simulate timeseries of available production provide some diversity because of the combo between the energetic mix and the climate vectors. Nevertheless, as the number of agents grows, the timeseries tend to exhibit similar patterns. This is particularly visible when creating a correlation graph $ G_{1}(\mathcal{A},E_{1},\omega_{1}) $ with the same kind of metric as \cite{Garas2008} or \cite{Onnela2004} ($ d_{ij} = 1 - \rho_{ij}^{2} $). Indeed, very well defined clusters appear in the corresponding $ \epsilon $-graph for any values of $ \epsilon $.

@@

However, as seems intuitively understandable, cliques of this graph tend to exhibit very good utility values. Such  structures contain ideed a link between every two nodes, meaning that the overall correlation is quite small. Obviously, the $ \epsilon $ parameter is indirectly responsible for the sizes of the cliques, if it is too low, the $\epsilon$-graph of $ G_{2} $ will not provide enough cliques, conversely, if $\epsilon $ is too high, we loose important information as the graph becomes very dense and cliques tend to overlap strongly. For large values of $ \epsilon $ and for non trivial number of agents, finding cliques can even becomes computationally repulsive. Despite being direct and simple, improving the sizes of the cliques by increasing $ \epsilon $ seems too brutal. Furthermore, as the utility function is also focused on maximizing $ P_{\phi}(S) $, it might be the case that some cliques benefit from additional agents even if they don't form a clique anymore.

@@

\[ \tau_{i}(S) = \dfrac{\mathcal{U}_{\phi,\ P_{\phi}^{MIN}}(S) - \mathcal{U}_{\phi,\ P_{\phi}^{MIN}}(S-\{i\}) }{\mathcal{U}_{\phi,\ P_{\phi}^{MIN}}(S)} \]

@@

At this point, we have three degrees of freedom, namely the reliability ($ \phi $), the required power to enter the market ($P_{\phi}^{MIN}$), and the prunning parameter ($\epsilon$). For conveniance, we introduce the number of desired coalitions ($ N_{COAL} $) and we fix $\epsilon $ as the smallest value such that the resulting $ \epsilon $-graph contains at least $ N_{COAL} $ non overlaping cliques. The overall utility depends now on the grid requirements and on the number of desired coalitions.

@@

As visible on the left plot of figure \ref{Fig2}, the parameters $\phi$ and $P_{\phi}^{MIN}$ shape the utility function such that, if $ \phi $ is close to zero, the reliability requirement is very high and only small values of $ P_{\phi}^{MIN}$ could potentially lead to valid coalitions (and positive utility values). Conversely, the higher $\phi$, the less constraints are imposed to the coalitions and valid ones can arise for a larger spectrum of $ P_{\phi}^{MIN}$. Obviously, the highest utility values are found for high $ \phi $, because coalitions are able then to announce higher contract values, yielding higer utilities.

@@

\begin{center}
\begin{figure}[b]
  \includegraphics[scale=0.8]{./figure5/coals.pdf}
  \caption{Evolution of the coalitions for different values of grid requirement $ P_{\phi}^{MIN} $. Blue dots represents clique percolation results while red dots stands for random process results.}
  \label{Fig3}
\end{figure}
\end{center}

@@

Looking at figure \ref{Fig3}, we can see first that our algorithm seems to make generally a better job at finding high valued coalitions, but also that the results seem far more robust against how the grid positions its requirements. The blue dots allowed to enter the market seem indeed to outnumber the red dots, especially when the grid requirements are neither to low or to high ($ P_{\phi}^{MIN} = 5 $ in figure \ref{Fig3}).

@@

When the grid requirements vary, for very low $ P_{\phi}^{MIN} $ (close to zero), all coalitions for both algorithms are able to enter the market, yielding an acceptance percentage of $ 100 \% $. But as $ P_{\phi}^{MIN} $ increases, we see the percentage of the random procedure quickly dropping while it stays constant for our algorithm. For $ P_{\phi}^{MIN} = 4 $, we see that only a little more than half of the coalitions for the random case are able to enter while all of them enters for the clique percolation algorithm. Finally, when the grid requirement becomes to high, the acceptance percentage of our algorithm tends slowly to zero (only partly shown in this plot for readability).

\begin{center}
\begin{figure}
\setlength{\tabcolsep}{0em}
\begin{tabular}{cc}
   \includegraphics[scale=0.45]{./figure7/uti.eps} &
   \includegraphics[scale=0.45]{./figure4/survival.eps} \\
\end{tabular}
   \caption{Percentage of coalitions able to enter the market for different values of grid requirement $ P_{MIN} $. Blue curve represents clique percolation results while red curve stands for random process results}
   \label{Fig4}
\end{figure}
\end{center}

@@

In this paper we collected meteorological traces for weather stations sampling a given territory. After discetization of the space, each zone receives the climate vector corresponding to its mother weather station. Prosumers are then distributed randomly between zones and each prosumer receives a random number of DER and loads as well as diverse consumption habits. Simulations are then runned and allow us to obtain, for each prosumer, the timeserie describing the instantaneous power that he could potentially be wanted to sell on the market. Nevertheless, as entering on the energy market is constrained by the grid operator, such that only sufficiently reliable and sufficiently productive units are allowed to enter, prosumers have usually no other choice than forming coalitions. We then introduced a framework such that the grid requirements take a precise meaning and such that coalitions announce contract values accordingly.

Through this paper, our interest was on how we could circumscribe groups of prosumers that can rise above the grid requirements. In this direction, we removed seasonal variations from the prosumers timeseries and proposed to organize them over a correlation graph with a reversed metric. A greedy clique percolation process on the $ \epsilon $-graph enables us to finally establish coalitions of agents. We showed in section \ref{sec:results} that our algorithm was actually finding coalitions with high utility values but also that it was far more robust than a random search over coalition structures when the grid operator increases his requirements (Figure \ref{Fig4}).

An interresting lead for future work would be the introduction of a payoff allocation towards the prosumers such that the stability against player defection could be analysed. Besides, we believe that not restricting the algorithm to non overlapping coalitions and studying the strategies and weights of nodes with multiple options could lead to interesting works.

@@

% use section* for acknowledgement
\section*{Acknowledgment}

The authors would like to thank...

>>>>

% author names and affiliations
% use a multiple column layout for up to three different
% affiliations
\author{\IEEEauthorblockN{Nicolas Gensollen}
\IEEEauthorblockA{CNRS SAMOVAR, \\
Institut Mines Telecom}
\and
\IEEEauthorblockN{Vincent Gauthier}
\IEEEauthorblockA{CNRS SAMOVAR, \\
Institut Mines Telecom}
\and
\IEEEauthorblockN{Michel Marot\\ and Monique Becker}
\IEEEauthorblockA{CNRS SAMOVAR, \\
Institut Mines Telecom}}

@@

\begin{abstract}
%\boldmath

In a smart grid environement, we study coalitions of prosumers (agents that produce but also consume depending on time) which aim at selling energy to the grid. We reckon that the grid is able to specify two requirements (stability and minimum production) that enable a coalition to enter the market. We then design an algorithmm that seeks for uncorrelated profiles among the prosumers in order to form coalitions that are statistically stable enough to pass the grid requirements. The proposed algorithm uses complex system tools such as correlation graphs or clique percolation, and remains simple enough such that coalitions are formed in reasonable time. We validate the algorithm against a random procedure and show that, it not only performs better in term of social welfare, but that it is also more robust against grid minimum production threshold variations.

%There is a growing interest for coalitional concepts in the smart grid, spanning multiple objectives such as power losses reduction, benefits maximization or market stability... In this paper, we study coalitions of prosumers (agents that produce but also consume depending on time) which aim at selling energy to the grid. We reckon that the grid is able to specify two requirements (stability and minimum production) that enable a coalition to enter the market. Based on probability distributions of units available power, coalitions which are able to achieve low standard deviations are considered as stable coalitions. As it seems intuitively understandable that maintaining stable coalitions would be less expensive in terms of communication load than highly versatible ones, our objective consists in forming coalitions that satisfy grid requirements both in terms of stability and minimum production. Our definition of stability being implicitely linked to correlation between agents timeseries (see section \ref{sec:model}), we organize them over a correlation graph and use a variation of clique percolation to construct valid coalitions in reasonable time.
\end{abstract}
% IEEEtran.cls defaults to using nonbold math in the Abstract.
% This preserves the distinction between vectors and scalars. However,
% if the conference you are submitting to favors bold math in the abstract,
% then you can use LaTeX's standard command \boldmath at the very start
% of the abstract to achieve this. Many IEEE journals/conferences frown on
% math in the abstract anyway.

@@

In this paper, our focus is on how improving statistically the production stability of coalitions of prosumers while decreasing the communication needs for staying in a stable state. The idea is that, by using prediction techniques, coalitions can propose contract values to the grid, enabling it to schedule production on its side accordingly. Actually, contract values are particularly relevant in power exchanges contexts where energy is traded based on predictions (day-ahead estimations for instance). It seems obvious that, in such systems, participants should try to minimize prediction errors in order to maintain a stable state of grid operation, and that some penalty rules should be applied to ensure that coalitions are willing to report contract values correctly. Furthermore, as the coalition productions are supposed to come from (uncertain) renewables, the notion of reliability comes into play with a visible tradeoff between contract values worth and reliability.

In this paper, we seek at optimizing this tradeoff, that is forming coalitions able to announce “high contract values with high reliability”. We thus aim at:

%In this paper, our concern is on how prosumers should be aggregated in order to optimize this tradeoff, that is, how coalitions should be formed with the intention of announcing “high contract values with high reliability”. We thus aim at:

@@

On a narrower scale, the authors of \cite{Kota2011} study the formation of virtual power plants (VPP) composed of multiple self-interested DER. On the grid side, two requirements for the formation of virtual power plants are considered : the reliability of supply and the minimization of entities the grid has to deal with. From this, \cite{Kota2011} builds a pricing mechanism that encourages VPP to report true estimates of their aggregated production and penalizes prediction errors. A redistribution scheme of the VPP to the DER is also constructed such that the payoff allocation lies in the core of the game, meaning that no DER has an incentive to leave the coalition.

@@

The setting is thus similar with some fiancial studies on stock exchanges, where researchers tried to find relevant clusters of stocks based on their daily prices variations. The problem of clustering is usually approached by means of a similarity measure and completed by a clustering technique such as K-means or hierarchical clustering. Nevertheless, in \cite{Mantegna1999} Mantegna introduced another approach where stocks, represented by the timeseries of their daily log returns, are organized in a graph such that stocks (vertices) exhibiting similar price fluctuation patterns are close to each other. This closeness notion is formalized with a similarity measure based on Pearson correlation coefficient ($ d_{ij} = \dfrac{1}{2}\sqrt{2(1-\rho_{ij})} $ or $ d_{ij} = 1 - \rho_{ij}^{2} $ ) that enables to weight the edges of the graph. Because there is a distance between every two nodes, \cite{Mantegna1999} obtains a complete graph where precious information is flooded. In order to discard unsignificant information, \cite{Mantegna1999} used a minimum spanning tree oriented technique and achieves a hierarchical clustering of the stocks. Since then, several other filtering techniques have been proposed, such as k-nearest graphs, or $ \epsilon $-graphs, that relax the tree assumption \cite{Onnela2004}.

@@

Presented in this way, the timeseries clustering task seems very close to graph community detection. Communities in networks are indeed often seen as groups of nodes exhibiting high internal densities of links as well as a low density across communities \cite{Newmanb}. Although several techniques exists (based on different graph properties : modularity \cite{Newmanb}, edge betweeness \cite{Girvan2002}, spectrum of Laplacian \cite{Newman}...), the clique percolation algorithm \cite{Lancichinetti} uses directly this observation and the idea behind it is actually quite simple. It starts indeed by searching for cliques (a complete subgraph) in the given graph as potential community cores. Because other nodes can naturally belong to a community with a less strong connectivity requirement, the algorithm uses a fitness function that quantifies the quality (according to some criterions) of a community structure inside the graph. Expansed cliques alternatively look in their respective neighborhood, select the node that yields the best increase in fitness, and finally incorporate it. The percolation algorithm stops when seeds cease to grow or when all nodes are affected to at least one seed. At this point, some coalitions may be partially overlapping, meaning that a node can belong to multiple communities. Detection of overlapping communities is actually a very active field of research, especially in social networks where persons often belongs to several communities (family, friends, colleagues...).

@@

A major concern while designing our model was its ability to translate correctly realistic patterns (consumption and production) as well as realistic geographical correlations between patterns. Largely because of these reasons (but also because they are easily, and sometimes freely, available online) we chose to use meteorological traces as inputs. In this paper we used namely both french data from 2006 to 2012 sampled every 3 hours \cite{Infoclimat} (similar data for the United States covering 2010 can also be found at \cite{NCDC}). The weather stations already offer a quite good sampling of a given territory, and, with the growth of personal weather stations constantly updating data bases, the mesh becomes finer and finer. As shown in the first part of figure 1, the first step consists in discretizing the studied zone around well chosen weather stations and gathering traces for these stations. In this paper, we consider that prosumers can only produce through wind turbines and photovoltaic panels (PV). On the other hand, we built their consumption patterns according to two major cycles :
\begin{itemize}
\item \textbf{Daily cycles} : Consumption is low during night, and higher during the day with two picks in the morning and evening. Some noise is added so that prosumers have similar but not identical cycles.
\item \textbf{Seasonal cycles} : Consumption is higher in the winter because of heating and low in the summer (air conditioning is not considered). Temperature traces were the principal ingredient for modeling these cycles.
\end{itemize}

@@

At this point, modeling agents consists in fixing a few parameters (most of the time drawn from random distributions) such as the geographical position, the number of wind turbines, PV, appliances, the temperature of confort, and so on... The objective is that, depending on the weather of his zone, the DER and appliances he owns, and the way he decides to heat his home, a given prosumer is able to compute his production/consumption at any time. We thus used simple and known power curve models for linking weather variables to output power, such that a wind turbine (or a PV for instance) takes a wind speed (a solar irradiance) as input and gives an instantaneous output power \cite{Kota2011} \cite{windturbinemodel}. Any agent is thus able to compute at any time how much he produces and consumes.

@@

The reliability stipulates that for any coalition S willing to join the market, the probability of S value (at any instant t) being below its contract value should at most be $ \phi $. That is, $ \forall S,\ \forall t,\ Pr[P_{S}(t) \leq P_{S}^{CRCT} ] \leq \phi $ (for consistency we restrict $ \phi $ to small values ( $ \phi << 0.5 $)). The minimum contract value under $ \phi $ states that only coalitions with higher contract values ($ P_{S}^{CRCT} \geq P_{\phi}^{MIN} $) will be accepted.

At this point, if a coalition S whishes to join the market, it has to choose a contract value that fulfill the two conditions. For simplification, we consider that coalitions will always apply the same economically consistent strategy of announcing the highest possible contract value that obeys the reliability rule (a value we denote by $ P_{\phi}(S) $). If $ P_{\phi}(S) \geq P_{\phi}^{MIN} $, meaning that it also obeys the minimum contract value rule, then S announces this value on the market : $ P_{S}^{CRCT} = P_{\phi}(S) $. Otherwise, coalition S is not able to enter.
Basically, a coalition S is valid if and only if :
\begin{equation}
\left\{ \begin{array}{lll}
		\forall t,\ Pr[ P_{S}(t) \leq P_{\phi}(S)] \leq \phi\ \textit{{\scriptsize (reliability rule)}} \\
		and\ P_{\phi}(S) \geq P_{\phi}^{MIN}\ \textit{{\scriptsize (min value rule)}}

\end{array} \right. 
\label{rules}
\end{equation}
We now choose a very simple utility function (eq. \ref{utility}) that derives directly from the above remarks. If a coalition cannot provide a valid contract value, it receives naturally a utility of zero. Furthermore, it seems obvious that the utility should increase with the contract value. The $ 1/|S| $ term in eq. \ref{utility} indicates that we favorise small coalitions, mainly because they are easier to maintain in terms of communications.

\begin{equation}
 \mathcal{U}_{\phi,\ P_{\phi}^{MIN}}(S) = \mathbf{1}_{\textit{S\ valid}} \dfrac{P_{\phi}(S)}{|S|} 
\label{utility}
\end{equation}

@@

\begin{equation}
\left\{ \begin{array}{lll}
		\mu_{ij} = \mu_{i} + \mu_{j} \\
		\sigma_{ij} = \sqrt{\sigma_{i}^{2} + \sigma_{j}^{2} + \rho_{ij}\sigma_{i}\sigma_{j}}
\end{array} \right.
\label{parameters}
\end{equation}

@@

\begin{equation}
\dfrac{1}{2} \left[ 1+ erf \left( \dfrac{P_{\phi}(ij) - \mu_{ij}}{\sigma_{ij}\sqrt{2}} \right) \right] \leq \phi
\label{reliability}
\end{equation}

The strategy of $ \{ij\} $ consists in maximizing its contract value as long as it respects the inequality \ref{reliability}, e.g to annonce $ P_{\phi}(ij) = \mu_{ij} - \sigma_{ij}\sqrt{2}erf^{-1}(1-2 \phi ) $. It thus appears (as it was intuitively understandable) that, for equivalent sizes, coalitions with low relative standard deviations ( $ \sigma_{ij} / \mu_{ij} $ ) are able to announce higher contract values.

@@

The model we used in order to simulate timeseries of available production provides some diversity because of the combination between the energetic mix and the climate vectors. Nevertheless, as the number of agents grows, the timeseries tend to exhibit similar patterns. This is apparent when creating a correlation graph $ G_{1}(\mathcal{A},E_{1},\omega_{1}) $ with the same kind of metric as \cite{Garas2008} or \cite{Onnela2004} ($ d_{ij} = 1 - \rho_{ij}^{2} $), where well defined clusters appear in the $ \epsilon $-graph for any values of $ \epsilon $.

@@

However, as seems intuitively understandable, cliques of this graph tend to exhibit very good utility values. Such  structures contain indeed a link between every two nodes, meaning that the overall correlation is quite small. Obviously, the $ \epsilon $ parameter is indirectly responsible for the sizes of the cliques, if it is too low, the $\epsilon$-graph of $ G_{2} $ will not provide enough cliques, conversely, if $\epsilon $ is too high, we loose important information as the graph becomes very dense and cliques tend to overlap strongly. For large values of $ \epsilon $ and for non trivial number of agents, finding cliques can even become computationally repulsive. Despite being direct and simple, improving the sizes of the cliques by increasing $ \epsilon $ seems too brutal. Furthermore, as the utility function is also focused on maximizing $ P_{\phi}(S) $, it might be the case that some cliques benefit from additional agents even if they don't form a clique anymore.

@@

\begin{equation}
\tau_{i}(S) = \dfrac{\mathcal{U}_{\phi,\ P_{\phi}^{MIN}}(S) - \mathcal{U}_{\phi,\ P_{\phi}^{MIN}}(S-\{i\}) }{\mathcal{U}_{\phi,\ P_{\phi}^{MIN}}(S)}
\label{tau}
\end{equation}

@@

At this point, we have three degrees of freedom : the reliability ($ \phi $), the required power to enter the market ($P_{\phi}^{MIN}$), and the prunning parameter ($\epsilon$). For convenience, we introduce the number of desired coalitions ($ N_{COAL} $) and we fix $\epsilon $ as the smallest value such that the $ \epsilon $-graph contains at least $ N_{COAL} $ non overlaping cliques. The overall utility depends now on the grid requirements and on the number of desired coalitions.

@@

As visible on the left plot of figure \ref{Fig2}, the parameters $\phi$ and $P_{\phi}^{MIN}$ shape the utility function such that, if $ \phi $ is close to zero, the reliability requirement is very high and only small values of $ P_{\phi}^{MIN}$ could potentially lead to valid coalitions (and positive utility values). Conversely, the higher $\phi$, the less constraints are imposed to the coalitions and valid ones can arise for a larger spectrum of $ P_{\phi}^{MIN}$. Obviously, the highest utility values are found for high $ \phi $, because coalitions are able then to announce higher contract values, yielding higher utilities.

@@

\begin{center}
\begin{figure}[b]
  \includegraphics[scale=0.8]{./figure5/coals.pdf}
  \caption{Evolution of the coalitions for different values of $ P_{\phi}^{MIN} $ for clique percolation (blue dots) and random process (red dots).}
  \label{Fig3}
\end{figure}
\end{center}

@@

Looking at figure \ref{Fig3}, we can see first that our algorithm seems to make generally a better job at finding high valued coalitions, but also that the results seem far more robust against how the grid positions its requirements. The blue dots allowed to enter the market seem indeed to outnumber the red dots, especially when the grid requirements are neither too low or too high ($ P_{\phi}^{MIN} = 5 $ in figure \ref{Fig3}).

@@

When the grid requirements vary, for very low $ P_{\phi}^{MIN} $ (close to zero), all coalitions for both algorithms are able to enter the market, yielding an acceptance percentage of $ 100 \% $. But as $ P_{\phi}^{MIN} $ increases, we see the percentage of the random procedure quickly dropping while it stays constant for our algorithm. For $ P_{\phi}^{MIN} = 4 $, we see that only a little more than half of the coalitions for the random case are able to enter while all of them enters for the clique percolation algorithm. Finally, when the grid requirement becomes too high, the acceptance percentage of our algorithm tends slowly to zero (only partly shown in this plot for readability).

\begin{center}
\begin{figure}
\setlength{\tabcolsep}{0em}
\begin{tabular}{cc}
   \includegraphics[scale=0.45]{./figure7/uti.eps} &
   \includegraphics[scale=0.45]{./figure4/survival.eps} \\
\end{tabular}
   \caption{Percentage of coalitions able to enter the market for different values of $ P_{MIN} $ for clique percolation (blue) and random process (red).}
   \label{Fig4}
\end{figure}
\end{center}

@@

We believe that the originality of this paper lies in its willingness to exploit the de-correlation of prosumer profiles in order to build stable coalitions. In this direction, we presented a model based on meteorological traces, that captures the complex "\textit{energetic mix/climate vectors}" combination and generates realistic production and consumption patterns. We then built a framework that enables the grid to specify stability and minimum production requirements for filtering the coalitions. On this basis, we proposed a simple algorithm that seeks for uncorrelated prosumer patterns as potential seeds and expand them in coalitions able to rise above the grid requirements.

The algorithm is validated against a random process, where we show that it performs better (coalitions are more stable and the overall production is more important) and that it exhibits a higher robustness/flexibility against grid requirements changes.

%In this paper we collected meteorological traces for weather stations sampling a given territory. After discetization of the space, each zone receives the climate vector corresponding to its mother weather station. Prosumers are then distributed randomly between zones and each prosumer receives a random number of DER and loads as well as diverse consumption habits. Simulations are then runned and allow us to obtain, for each prosumer, the timeserie describing the instantaneous power that he could potentially be wanted to sell on the market. Nevertheless, as entering on the energy market is constrained by the grid operator, such that only sufficiently reliable and sufficiently productive units are allowed to enter, prosumers have usually no other choice than forming coalitions. We then introduced a framework such that the grid requirements take a precise meaning and such that coalitions announce contract values accordingly.

%Through this paper, our interest was on how we could circumscribe groups of prosumers that can rise above the grid requirements. In this direction, we removed seasonal variations from the prosumers timeseries and proposed to organize them over a correlation graph with a reversed metric. A greedy clique percolation process on the $ \epsilon $-graph enables us to finally establish coalitions of agents. We showed in section \ref{sec:results} that our algorithm was actually finding coalitions with high utility values but also that it was far more robust than a random search over coalition structures when the grid operator increases his requirements (Figure \ref{Fig4}).

Interresting leads for future works would be the use of correlated clusters to reduce the number of entities the algorithm has to deal with, or the introduction of a payoff allocation towards the prosumers such that the stability against player defection could be analysed through game theory. Showing that maintaining the coalitions formed with our algorithm necessitates less communication and less storage capacity could also conduct to a stimulating project. Besides, we believe that not restricting the algorithm to non overlapping coalitions and studying the strategies and weights of nodes with multiple options could lead to interesting works.

@@
